Composable Concurrency Models

### 2016-12-01 paper/paper.tex
Made some final tweaks

Processes themselves are often composed of multiple concurrent threads of execution. A processor can only execute one thread at a time, so multitasking is accomplished by rapidly switching between threads~\cite{Liu1973}. Although concurrent threads may appear to be executed simultaneously, truly parallel execution can only take place across multiple processors.

@@

Consider a third example (Figure~\ref{figure:time}) in which two threads $A$ and $B$ interact with a variable $x$ noninstantanously. Invocation is denoted by the leftmost point of an event and completion by the rightmost point. The dotted line indicates the instant at which the operation takes effect. The lower thread invokes a read while the value of $x$ is set to $2$. The upper thread invokes and completes a write between the time that the read is invoked and the value is actually read. The read operation then returns $3$ even though $x$ was equal to $2$ at the time it was invoked.

@@

Figure~\ref{figure:serializability} illustrates a set of serializable read and write operations. Even though the executions of each operation occur out of order and even intersect, the events themselves can be arranged serially.

@@

Atomic variables are used to share independent objects that do not demand coordinated updates but need to be shared by multiple threads~\cite{Swalens2014}. Atomic variables can only be read and mutated by operations that are guaranteed to be linearizable. For example, the atomic operation \code{compare-and-swap} compares the current value of a variable to a given value and only performs a write if the values are equal~\cite{Swalens2014}. \code{compare-and-swap} ensures that a new value cannot be written based on outdated information: Suppose that a thread $A$ reads a variable $x$ and begins computing a new value. At the same time, another thread $B$ modifies $x$. When $A$ tries to set the value of $x$ via \code{compare-and-swap}, the write will fail. As a result, all writes on $x$ are guaranteed to be linearizable.

@@

Atomic variables often serve as the basis for higher-level concurrency control mechanisms such as semaphores and locks, which in turn support the implementation of concurrent data structures. A counting semaphore is an atomic variable that serves as a record of how many units of a shared resource, such as a connection or a variable, are available. By taking advantage of atomic operations, the semaphore can be safely updated and will reliably determine whether a shared resource can be used. A single thread can then test if a resource is available and acquire it before proceeding, thereby preventing race conditions~\cite{Swalens2014}. A lock, also referred to as a mutex, guarantees mutual exclusion, which requires that two threads cannot operate on a shared object at the same time. Locks are commonly implemented by a binary semaphore (that is, a semaphore that simply indicates whether a single resource is available).

@@

Conceptually, STM simplifies concurrency because it allows a transaction to be thought of as a single strictly serializable operation. A thread cannot observe changes made to other threads while a transaction is in progress, nor can other threads observe modifications by that thread until the transaction is completed. Only when a transaction completes successfully will changes will become visible to other threads. Wrapping a set of operations within a transaction eliminates the need to secure locks, leading to more simple programs. Practically, STM is useful in any situation in which a shared object needs to be accessed and modified by multiple threads~\cite{Swalens2014}.

@@

Other concurrently models avoid the use of shared memory by restricting threads to private memory. Threads communicate strictly by message passing, which avoids issues such as race conditions. Sending messages and reacting to received messages are inherently asynchronous processes, so communicating threads have long been a solution to describing concurrent operations. Limiting thread communication to message passing also enforces modularity and limits possible negative side effects. A wide variety of communicating thread models exist and are often used to implement event loops or handle communication with external systems~\cite{Swalens2014}.

@@

Other models, most notably the actor model, rely on asynchronous message passing to specific entities (actors) rather than named channels~\cite{Agha1986}. Actors can only communicate by asynchronous message passing. Upon receiving a message, an actor can choose to perform an operation on a resource that it controls, send messages to other actors, create new actors, determine the behavior used for the next received message, or ignore the message entirely.

@@

Proxies are a general term used to describe placeholders for values that are the result of some concurrently executed operation. Unlike communicating thread models, which rely on message passing between dedicated threads, a proxy executes some function in a new thread and delivers the result upon completion~\cite{Swalens2014}.

@@

In practice, proxies are used to execute long-running operations such as rendering or network requests~\cite{Swalens2014} without blocking other operations. Promise objects, which can be easily passed and composed, offer more flexibility than constructs such as asynchronous callbacks.

@@

Even if concurrency models guarantee their own correctness, issues may arise when different models are used together. For example, an implementation of an STM that guarantees linearizability assumes that all shared resources are managed by the STM~\cite{Shavit1995}. If such a resource is modified outside of the STM, an unexpected interleaving of operations could result.

Swalens et al.~\cite{Swalens2014} attempted to identify issues that arise when different concurrency models are combined, hoping that future work could identify common ``building blocks'' that could be used to compose a variety of concurrency models. The study surveyed all of the concurrency models available in the Clojure programming language and used them in within one another to see what types of correctness issues were encountered.

@@

Informally, safety guarantees that ``nothing bad will happen.'' In other words, given a correct input, a program will not produce an incorrect result~\cite{Swalens2014}. This property is also referred to as partial correctness. Safety is generally achieved in the management of shared resources. STM, for instance, only allows shared memory to be accessed through transactions; communicating threads only allow data to be shared through message passing.

@@

\begin{itemize}
  \item A model might reexecute code containing another concurrency model that performs an irrevocable action. For instance, if a transaction were to contain an operation that sent a message on a channel, reexecutions of the transaction would cause the message to be sent repeatedly.

@@

  \item A model might not guarantee safety or liveness by design, making composability impossible. Perhaps obviously, a model that does not guarantee safety or liveness would violate the safety guarantee of an outer model unless such a conflict was explicitly prevented.
\end{itemize}

The study noted that some types of bad interactions are prevented by Clojure; for instance, sending a message to an agent from within a transaction is delayed until the transaction succeeds. The study also noted that similar safeguards could prevent other negative interactions. However, preventing every type of negative interaction between models might undermine the purpose of the models to begin with; channels, as mentioned previously, block by design and are inherently incompatible with other models. With that in mind, further research would likely seek to decompose concurrency models into common elements and provide a method to compose those elements safely and efficiently.

@@

Marr and D'Hondt~\cite{Marr2012} surveyed a variety of concurrent and parallel programming concepts such as immutability, critical sections, ownership, and common concurrency models to identify a unifying concurrency abstraction that could be implemented in a high-level language virtual machine. A unifying abstraction would enable each concurrency model in a high-level language to be represented in machine code without sacrificing performance or semantics. For instance, a concurrency model in Clojure could be accurately represented by an underlying abstraction in the Java Virtual Machine, and code written in the Java language could interact with that model correctly. To identify the concepts that would need to be supported at the virtual machine level, Marr and D'Hondt selected four questions:

@@

  \item Does the concept require runtime support to guarantee its semantics? Semantic guarantees may be enforced by a compiler but not the virtual machine. For instance, immutable objects in one language may not be immutable in machine code, and therefore may be mutated by another language.

@@

From those questions, 26 concepts were identified that would benefit from better low-level support. Of those, 18 suffered from loss of semantic guarantees when compiled to machine code. With that in mind, Marr and D'Hondt derived the following requirements for a unifying abstraction:

\begin{itemize}
  \item Managed mutation and execution: A model may impose rules on how objects can be modified; therefore, mutation must be handled in a way that allows those rules to be enforced. Similarly, a model might restrict the invocation of operations on objects, and those rules must be enforced. For example, a thread should not be permitted to mutate a variable that a model declares immutable.

  \item Ownership: All mutation an execution on an object is regulated relative to some ``owning'' entity, so ownership should be supported in a way that allows for adaptable mutation and execution rules.

@@

\begin{figure}[ht]
  \centering
  \resizebox{0.55\linewidth}{!}{\input{../figures/obmop.tikz}}
  \caption{Marr and D'Hondt's object-based meta-object protocol~\cite{Marr2012}.}
\label{figure:obmop}
\end{figure}

@@

To demonstrate the suitability of the MOP as a low-level abstraction, Marr and D'Hondt implemented several concurrency models. The MOP was able to support LRSTM (a STM for the Smalltalk language), Clojure's agents (which rely on communicating threads), active objects (a variation on proxies), and several other models. While the MOP was able to successfully enforce the semantics of each model, the resulting performance cost suggested that the MOP must be implemented at the virtual machine level to sufficiently reduce the abstraction overhead. Marr and D'Hondt also note that regarding the owner of an object as the only entity able to restrict interaction is somewhat limiting. For instance, the behavior of the system when different semantics interact needs to be explicitly specified.

>>>>

Processes themselves are often composed of multiple concurrent threads of execution that each work on a distinct task. A processor can only execute one thread at a time, so multitasking is accomplished by rapidly switching between threads~\cite{Liu1973}. Although concurrent threads may appear to be executed simultaneously, truly parallel execution can only take place across multiple processors.

@@

Consider a third example (Figure~\ref{figure:time}) in which two threads $A$ and $B$ interact with a variable $x$ noninstantanously. Invocation is denoted by the leftmost point of an event and completion by the rightmost point. The dotted line indicates the instant at which the operation takes effect. Thread $B$ invokes a read while the value of $x$ is set to $2$. Thread $A$ invokes and completes a write between the time that the read is invoked and the value is actually read. The read operation then returns $3$ even though $x$ was equal to $2$ at the time it was invoked.

@@

Figure~\ref{figure:serializability} illustrates a set of serializable read and write operations. Even though the executions of each operation occur out of order and even intersect, the events themselves can be arranged serially; no side effects are introduced by the fact that they are executed concurrently.

@@

Atomic variables are used to share independent objects that do not demand coordinated updates but need to be shared by multiple threads~\cite{Swalens2014}. Atomic variables can only be read and mutated by operations that are guaranteed to be linearizable. For example, the atomic operation \code{compare-and-swap} compares the current value of a variable to a given value and only performs a write if the values are equal~\cite{Swalens2014}. \code{compare-and-swap} ensures that a new value based on outdated information cannot be written: Suppose that a thread $A$ reads a variable $x$ and begins computing a new value. At the same time, another thread $B$ modifies $x$. When $A$ tries to set the value of $x$ via \code{compare-and-swap}, the write will be denied.

@@

Atomic variables often serve as the basis for higher-level concurrency control mechanisms such as semaphores and locks, which in turn support the implementation of concurrent data structures. A counting semaphore is an atomic variable that serves as a record of how many units of a shared resource, such as a connection or a pooled thread, are available. By taking advantage of atomic operations, the semaphore can be safely updated and will reliably determine whether a shared resource can be used. A single thread can then test if a resource is available and acquire it before proceeding, thereby preventing race conditions~\cite{Swalens2014}. A lock, also referred to as a mutex, guarantees mutual exclusion, which requires that two threads cannot operate on a shared object at the same time. Locks are commonly implemented by a binary semaphore (that is, a semaphore that simply indicates whether a single resource is available).

@@

Conceptually, STM simplifies concurrency because it allows a transaction to be thought of as a single strictly serializable operation. A thread cannot observe changes made to other threads while a transaction is in progress, nor can other threads observe modifications by that thread until the transaction has completed successfully. Wrapping a set of operations within a transaction eliminates the need to secure locks, leading to more simple programs. A transaction that modifies several shared variables, for example, would not have to secure locks for all of them. Practically, STM is useful in any situation in which multiple shared objects need to be accessed and modified by multiple threads~\cite{Swalens2014}.

@@

Other concurrently models avoid the use of shared memory by restricting threads to private memory. Threads communicate strictly by message passing, which avoids issues such as race conditions. Because of the higher level of isolation, communicating threads have long been a solution to describing concurrent operations. Limiting thread communication to message passing also enforces modularity and limits possible negative side effects. A wide variety of communicating thread models exist and are often used to implement event loops or handle communication with external systems~\cite{Swalens2014}.

@@

Other models, most notably the actor model, rely on asynchronous message passing to specific entities (actors) rather than named channels~\cite{Agha1986}. Upon receiving a message, an actor can choose to perform an operation on a resource that it controls, send messages to other actors, create new actors, determine the behavior used for the next received message, or ignore the message entirely.

@@

Proxies are a general term used to describe placeholders for values that are the result of some concurrently executed operation. Unlike communicating thread models, which rely on message passing between dedicated threads, a proxy executes some task in a new thread and delivers the result upon completion~\cite{Swalens2014}.

@@

In practice, proxies are used to execute long-running operations such as rendering or network requests~\cite{Swalens2014} without blocking other operations. For example, to eliminate the need for a program to block on a long network request, a promise could be used to complete the network request in another thread and specify an action that should take place once the result becomes available. Promise objects, which can be easily passed, composed, and chained, offer more flexibility than constructs such as asynchronous callbacks. In fact, many mainstream languages now directly support proxies.

@@

Even if a concurrency model guarantees correctness when used alone, issues may arise when it is used with other models. For example, an implementation of an STM that guarantees linearizability assumes that all shared resources are managed by the STM~\cite{Shavit1995}. If such a resource is modified outside of the STM, an unexpected interleaving of operations could result.

Swalens et al.~\cite{Swalens2014} attempted to identify issues that arise when different concurrency models are combined, hoping that future work could identify common ``building blocks'' that could be used to compose a variety of concurrency models. The study surveyed all of the concurrency models available in the Clojure programming language and used them within one another to see what types of correctness issues were encountered.

@@

Informally, safety guarantees that ``nothing bad will happen.'' In other words, given a correct input, a program will not produce an incorrect result~\cite{Swalens2014}. This property is also referred to as partial correctness. In the context of concurrent programming, safety is generally achieved by the management of shared resources. STM, for instance, only allows shared memory to be accessed through transactions; communicating threads only allow data to be shared through message passing.

@@

\begin{itemize}
  \item A model might reexecute code containing another concurrency model that performs an irrevocable action. If a block of code such as a transaction were to contain an operation that sent a message on a channel, reexecutions of that block would cause the message to be sent repeatedly.

@@

  \item A model might not guarantee safety or liveness by design, making safe or live composability impossible.
\end{itemize}

The study noted that some types of bad interactions are prevented by Clojure; as an example, sending a message to an agent from within a transaction is delayed until the transaction succeeds. The study also noted that similar safeguards could prevent other negative interactions. However, preventing every type of negative interaction between models might undermine the purpose of the models to begin with; channels, as mentioned previously, block by design and are inherently incompatible with other models. With that in mind, further research would likely seek to decompose concurrency models into common elements and provide a method to compose those elements safely and efficiently.

@@

Marr and D'Hondt~\cite{Marr2012} surveyed a variety of concurrent and parallel programming concepts such as immutability, critical sections, ownership, and common concurrency models to identify a unifying concurrency abstraction that could be implemented in a high-level language virtual machine. A unifying abstraction would enable each concurrency model in a high-level language to be represented in machine code without sacrificing performance or semantics. For instance, a concurrency model in Clojure could be accurately represented by an underlying abstraction in the Java Virtual Machine, and code written in the Java programming language could interact with that model correctly. To identify the concepts that would need to be supported at the virtual machine level, Marr and D'Hondt selected four questions:

@@

  \item Does the concept require runtime support to guarantee its semantics? Semantic guarantees may be enforced by a compiler but not the virtual machine. For example, immutable objects in one language may not be immutable in machine code, and therefore may be mutated by another language.

@@

From those questions, 26 concepts were identified that would benefit from runtime support. Of those, 18 suffered from loss of semantic guarantees when compiled to machine code; the survey mentioned transactions as an example. With those concepts in mind, Marr and D'Hondt derived the following requirements for a unifying abstraction:

\begin{itemize}
  \item Managed mutation and execution: A model may impose rules on how objects can be modified; therefore, mutation must be handled in a way that allows those rules to be enforced. By way of example, a thread should not be permitted to mutate a variable that a model declares immutable. Similarly, a model may restrict the invocation of operations on objects.

  \item Ownership: All mutation and execution on an object is regulated relative to some ``owning'' entity, so ownership should be supported in a way that allows for adaptable mutation and execution rules.

@@

\begin{figure}[ht]
  \centering
  \resizebox{0.55\linewidth}{!}{\input{../figures/obmop.tikz}}
  \caption{Marr and D'Hondt's owenership-based meta-object protocol~\cite{Marr2012}.}
\label{figure:obmop}
\end{figure}

@@

To demonstrate the suitability of the MOP as a low-level abstraction, Marr and D'Hondt implemented several concurrency models. The MOP was able to support LRSTM (a STM for the Smalltalk language), Clojure's agents (which rely on communicating threads), active objects (a variation on proxies), and several other models. While the MOP was able to successfully enforce the semantics of each model, the resulting performance cost suggested that the MOP must be implemented at the virtual machine level to sufficiently reduce the overhead of abstraction. Marr and D'Hondt also note that regarding the owner of an object as the only entity able to restrict interaction is somewhat limiting. For instance, the behavior of the system when different semantics interact needs to be explicitly specified.
